Application of Micro-Rotors to Induce Crystal Nucleation: Challenges, Takeaways, and Future Prospects\\ PhD Thesis

### 2025-07-04 Latex_files/Thesis/chapters/chapter 5/main.tex
fleshing out QPD chapter and edits to new chp 3&6

\section{Overview}
In this chapter, we begin with a brief discussion of the working 
principle behind a position detection device. We then provide a 
framework for replicating the scattering signal recorded by a 
quadrant photo diode. We show that extending the framework to 
capture information from an arbitrary shaped particle is simple 
in principle but difficult to interpret, we demonstrate as such 
by comparing the expected signal from a trapped sphere, to that 
of a trapped dimer.

@@

Beyond merely simulating the forces experienced by a spherical 
aggregate we also developed a simulative quadrant photo diode 
to replicate the results from a typical calibration test. This 
builds upon the work from \cite{Rohrbach2002} which applied 
Lorenz-Mie theory to replicate the response signal of a QPD 
being used in back focal-plane interferometry. Rather than be 
constrained to individual spheres we can now consider the 
expected response from any type of spherical aggregate. The
goal of such simulation software would be to simulate the 
expected experimental results from a complex particle whose 
exact shape and size may be difficult to determine.

In order to simulate a typical experimental set up with a QPD 
installed as a position detection system we need to evaluate 
the total magnitude of the electric field incident on the 
photo-diode surface. While trapping a micro-particle, the 
scattered and incident fields combine together and interfere 
with one another. These fields are collected by a condenser 
lens in the far field limit and are focused onto the QPD 
surface, the total intensity can be evaluated as:
\begin{align}
	I(x,y) = \epsilon_0c\left|
	\begin{bmatrix} 
		E_{i,x}(x,y)+E_{s,x}(x,y) \\ 
		E_{i,y}(x,y)+E_{s,y}(x,y) \\ 
		E_{i,z}(x,y)+E_{s,z}(x,y)
	\end{bmatrix} \right|^2 \times step(\theta_{NA}-\theta(x,y))
\end{align}

The last term is simply a representative step term that defines 
the outer limit by which we evaluate the electric field, this is 
analogous to our condenser lens removing noise from other light 
sources by only accepting light at a specific acceptance angle 
defined by its numerical aperture $NA_c$. Depending on the 
relative size of our particle we can adjust the acceptance angle, 
this has very little effect on the transverse signals, but for 
axial evaluations of a trapped particle the numerical aperture 
should be tuned so that the resultant response curve has negative 
slope in order to allow for axial position detection, the method 
for finding this angle $\theta_\Theta$ is discussed in 
\cite{Friedrich2012}.

The incident beam is simple enough to define given our set up 
parameters, for the sake of simplicity we assume that our beam is a Laguerre-Gaussian 
beam of mode $[0.0, 0.0]$ (which is simply a pure Gaussian beam). 
\textit{Ott} uses a point matching approach to approximate the beam shape 
coefficients of the incident field by fitting it to the far field estimate. 
From the QPD's perspective it is receiving light both from the incident 
and scattered beam simultaneously, as such both fields must be expressed
using outgoing vector spherical harmonics.
\begin{align}
	E_{\rm inc}(r)&=
	E_0\sum^\infty_{n}\sum^n_{m=-n}\left[a_{mn}{\bf M}^{(2)}_{nm}({\bf r})+
	b_{nm}{\bf N}^{(2)}_{nm}({\bf r}) \right] 
	\\
	E_{\rm scat}(r)&=E_0 \sum^\infty_n \sum^n_{m=-n} \left[
	p_{nm}{\bf N}_{nm}^{(2)}({\bf r})+q_{nm}{\bf M}_{nm}^{(2)}({\bf r})
	\right] 
\end{align}

@@

Where the denominator is the total intensity on the QPD while there is 
no particle within the trap. You would expect that $S_x$ and $S_y$ are
near identical for equal particle displacements, however due to the 
polarisation of the beam there will be a slight bias to the signals 
for motion along the direction of the polarisation vector. This is 
generally not an issue if the particle trajectory is sampled for a 
long enough time period. By converting from signal units to length
units (see \eqref{eq:conversion_factor}) the trap shape can be 
discerned from the QPD signals, assuming the fluid properties are
known to a high degree of accuracy.

@@

The discrepancy between our simulated QPD and the results 
from Rohrbach can be attributed to the fact that the position 
to signal error grows as you move further from trap focus. 
As the particle moves further from the trap focus the error 
in the attributed signal grows substantially \cite{Rohrbach2002}.
In most optical trapping experiments we usually are calibrating 
a strong trap where the mean square displacement falls well within 
the linear regime shown in fig.~\ref{fig:Rohrbach}. In which case 
the error between Rohrbach's model and our own is inconsequential 
for calibration purposes. With this any trajectory can be collected 
from the QPD by displacing and rotating the beam accordingly, 
later in chapters \ref{chapter:langevin_dynamics} and \ref{chapter:simulated_detection} we discuss the limits of 
accuracy that can be achieved using back focal plane interferometry 
to characterise non-spherical targets.

\subsection{Position detection}
\label{sec:position_detection}
In order to accurately capture the dynamics of a trapped particle, a 
position detection system is required. There are 3 possible methods 
of position detection: video-analysis, lateral-effect position sensing, 
and photodiodes.

Video analysis is ideally suited for multiple traps or situations where 
precision is not the top priority. Whereas lateral-effect and photodiode 
position detection are two examples of back-focal plane interferometry, 
where the interference pattern produced by the target particle is 
extrapolated to determine its position. In order for video analysis to 
match the accuracy of back-focal plane interferometry methods requires 
the camera's frame rate to exceed $1\ kHz$ which can be difficult to 
achieve while maintaining a decent resolution \cite{Gibson2008}. In 
comparison off the shelf back-focal plane detectors can achieve temporal 
resolutions anywhere from $10-100\ kHz$ \cite{BergSoerensen2004}.

A lateral-effect sensor has a similar output but works using a the 
entire sensor as a single cell analogous to the focal plane of the 
trapping beam. The four corners of the sensor act as anodes connected 
to a base plate cathode, as the beam moves across the surface of the 
detector each anode will experience a different photocurrent depending 
on how close the centre of the interference pattern is to each anode. 
The advantage of a lateral effect detector is that the linear regime 
is much larger than a QPD making it much better for monitoring the 
position of a trapped particle. However, Lateral-effect sensors are 
often limited in their spacial resolution due to high signal-to-noise 
ratios, requiring a high intensity of light on the sensor in order to 
get a clean signal. As a result, most optical force measurements are 
conducted using a QPD as opposed to a lateral-effect sensor, as often 
the displacement is small enough that the signal-displacement curve 
can be considered linear.
\begin{figure}[h!]
	\centering
	\includegraphics[height=\linewidth, angle=270]{QPD_Lateral_effect.pdf}
	\caption{Comparison between QPD and Lateral effect photodiodes.
		The four quadrants of a QPD (left) experience different photocurrents
		based on the total intensity of light incident on each section 
		(labelled A, B, C, D). 
		Whereas a Lateral effect sensor (right) uses the resistive properties
		of the photodiode surface to vary the create different photocurrents 
		passing through the anodes A, B, C, and D.}
\end{figure}

\section{Characterisation of asymmetric dimer dynamics via PSD analysis}
As discussed in \ref{sec:simulated_QPD}, one of the methods 
developed to work in conjunction with \cite{Vigilante2020} 
is a simulated quadrant photo diode for as a position 
detection system. While it is possible to extract all of 
the relevant dynamics from a simulation, from an experimental
standpoint characterising those dynamics is dependent on what
experimental techniques are used. Translational motion can be
characterised via a position detection system but angular 
motion is far more difficult to detect, let along characterise.

>>>>

@@

As demonstrated in Chapter 3, the motion and behaviour of a dimer 
is significantly different compared to that of a traditional sphere.
While this is simple enough to demonstrate in a simulation, where the 
position and orientation can be computed using the finite differences
method, from an experimental perspective these details can go unnoticed.
In order to accurately capture the dynamics of a trapped sphere, a 
position detection system is typically used. This can capture the 
translational motion of a trapped particle but relating the signals
produced to the rotational motion is not as straight forward.

The relationship between particle translations and light scattering 
was first demonstrated by Rohrbach, where they showed a linear change 
in the scattered light when the particle was displaced \cite{Rohrbach2002}. They further showed that the optical force exerted by the trapping 
beam could be determined from the change in scattered light \cite{Rohrbach2002}. Rohrbach's model would later be used to develop 
modern position detection systems, where by calibrating the scattering 
signal using one of the many calibration techniques one could achieve a highly accurate estimation of the translational motion in the $X$-$Y$ 
plane.The most common of which being the Power Spectral Density method, 
one of the greatest advantages of the PSD method is that periodic 
rotational motion is also captured in the power spectrum. As an 
example consider an elongated cylindrical particle that is wobbling
around the $Z$-axis at some periodic frequency 'f'. A power spectrum 
of that trajectory will show a peak at that frequency. If however the 
wobble is periodic but instead stochastic, then it is impossible to 
separate out the rotational motion from the translational motion. This
poses some challenges to experimental modelling of non-spherical 
particles, if we cannot glean any information on the orientation
of the trapped particle we cannot accurately distinguish between
sphere's and dimers

As a solution we developed a model similar to that of Rohrbach's but 
now utilising a \textit{mstm} to provide the light scattering for a 
dimer. We begin with a brief discussion of the working principle 
behind a position detection device. We then provide a framework for replicating the scattering signal recorded by a quadrant photo diode. 
We show that extending the framework to capture information from an 
arbitrary shaped particle is simple in principle but difficult to 
interpret, we demonstrate as such by comparing the expected signal 
from a trapped sphere, to that of a trapped dimer.

\subsection{Position Detection methods}
There are 3 common methods of position detection: video-analysis, lateral-effect position sensing, and photodiodes. The former is 
relies purely on cameras to track particle displacements; whereas 
the latter two are examples of focal-plane interferometry. 
Interferometry utilises the interference between the scattered and un-scattered light to measure displacement. Here we 
provide a brief outline of their applications, and limitations.

Video analysis is ideally suited for multiple traps or situations where 
precision is not the top priority. In order for video analysis to 
match the accuracy of back-focal plane interferometry methods requires 
the camera's frame rate to exceed $1\ kHz$ which can be difficult to 
achieve while maintaining a decent resolution \cite{Gibson2008}. In 
comparison off the shelf back-focal plane detectors can achieve temporal 
resolutions anywhere from $10-100\ kHz$ \cite{BergSoerensen2004}.

\begin{figure}[h!]
	\centering
	\includegraphics[height=\linewidth, angle=270]{QPD_Lateral_effect.pdf}
	\caption{Comparison between QPD and Lateral effect photodiodes.
		The four quadrants of a QPD (left) experience different photocurrents
		based on the total intensity of light incident on each section 
		(labelled A, B, C, D). 
		Whereas a Lateral effect sensor (right) uses the resistive properties
		of the photodiode surface to vary the create different photocurrents 
		passing through the anodes A, B, C, and D.}
\end{figure}

A lateral-effect sensor works using a single cell analogous to the focal plane of the trapping beam. The four corners of the sensor act as anodes connected to a base plate cathode, as the beam moves across the surface of the detector each anode will experience a different photocurrent depending 
on how close the centre of the interference pattern is to each anode. 
The advantage of a lateral effect detector is that the linear regime 
is much larger than a QPD making it much better for monitoring the 
position of a trapped particle. However, Lateral-effect sensors are 
often limited in their spacial resolution due to high signal-to-noise 
ratios, requiring a high intensity of light on the sensor in order to 
get a clean signal. As a result, most optical force measurements are 
conducted using a QPD as opposed to a lateral-effect sensor, as often 
the displacement is small enough that the signal-displacement curve 
can be considered linear.

The above informs how one can - to their best ability - measure small 
rotational movements in a trajectory. Relying on video analysis is 
unreliable as the frame rate required to accurately measure the 
orientation results in a loss of resolution. Likewise, simply measuring
the total intensity provides no means of separating out rotational and 
translational motion. Similar to Chapter 4, we need to utilise contrasting signals to estimate the change in orientation, and see if at all it is possible to distinguish between translational and rotational motion.

\section{Modelling the QPD response signal of a non-spherical particle}
In order to simulate a typical experimental set up with a QPD 
installed as a position detection system we need to evaluate 
the total magnitude of the electric field incident on the 
photo-diode surface. While trapping a micro-particle, the 
scattered and incident fields combine together and interfere 
with one another. These fields are collected by a condenser 
lens in the far field limit and are focused onto the QPD 
surface, the total intensity can be evaluated as:
\begin{align}
	I(x,y) = \epsilon_0c\left|
	\begin{bmatrix} 
		E_{i,x}(x,y)+E_{s,x}(x,y) \\ 
		E_{i,y}(x,y)+E_{s,y}(x,y) \\ 
		E_{i,z}(x,y)+E_{s,z}(x,y)
	\end{bmatrix} \right|^2 \times step(\theta_{NA}-\theta(x,y))
\end{align}

The last term is simply a representative step term that defines 
the outer limit by which we evaluate the electric field, this is 
analogous to a condenser lens removing noise from other light 
sources by only accepting light at a specific acceptance angle 
defined by its numerical aperture $NA_c$. Depending on the 
relative size of our particle we can adjust the acceptance angle, 
this has very little effect on the transverse signals, but for 
axial evaluations of a trapped particle the numerical aperture 
should be tuned so that the resultant response curve has negative 
slope in order to allow for axial position detection, the method 
for finding this angle $\theta_\Theta$ is discussed in 
\cite{Friedrich2012}.

The incident beam is simple enough to define given our set up 
parameters, for the sake of simplicity we assume that our beam is a 
Gaussian beam. \textit{Ott} uses a point matching approach to 
approximate the beam shape coefficients of the incident field by 
fitting it to the far field estimate. From the QPD's perspective 
it is receiving light both from the incident and scattered beam simultaneously, as such both fields must be expressed using outgoing 
vector spherical harmonics.
\begin{align}
	E_{\rm inc}(r)&=
	E_0\sum^\infty_{n}\sum^n_{m=-n}\left[a_{mn}{\bf M}^{(2)}_{nm}({\bf r})+
	b_{nm}{\bf N}^{(2)}_{nm}({\bf r}) \right] 
	\\
	E_{\rm scat}(r)&=E_0 \sum^\infty_n \sum^n_{m=-n} \left[
	p_{nm}{\bf N}_{nm}^{(2)}({\bf r})+q_{nm}{\bf M}_{nm}^{(2)}({\bf r})
	\right] 
\end{align}

@@

Where the denominator is the total intensity on the QPD while there is 
no particle within the trap. You would expect that $S_x$ and $S_y$ are
near identical for equal particle displacements, however due to the 
polarisation of the beam there will be a slight bias to the signals 
for motion along the direction of the polarisation vector. This is 
generally not an issue if the particle trajectory is sampled for a 
long enough time period. By converting from signal units to length
units (see Eq.\eqref{eq:conversion_factor}) the trap shape can be 
discerned from the QPD signals, assuming the fluid properties are
known to a high degree of accuracy.

@@

The discrepancy between our simulated QPD and the results 
from Rohrbach can be attributed to the fact that the position 
to signal error grows as you move further from trap focus. 
As the particle moves further from the trap focus the error 
in the attributed signal grows substantially \cite{Rohrbach2002}.
In most optical trapping experiments we usually are calibrating 
a strong trap where the mean square displacement falls well within 
the linear regime shown in fig.~\ref{fig:Rohrbach}. In which case 
the error between Rohrbach's model and our own is inconsequential 
for calibration purposes.

\section{Characterisation of asymmetric dimer dynamics via PSD analysis}
As discussed in Sec.~\ref{sec:simulated_QPD}, one of the methods 
developed to work in conjunction with \cite{Vigilante2020} 
is a simulated quadrant photo diode for as a position 
detection system. While it is possible to extract all of 
the relevant dynamics from a simulation, from an experimental
standpoint characterising those dynamics is dependent on what
experimental techniques are used. Translational motion can be
characterised via a position detection system but angular 
motion is far more difficult to detect, let along characterise.
